---
title: "Biostat 203B Homework 4"
subtitle: "Due Mar 9 @ 11:59PM"
author: "Nathaniel Boyle and 006525153"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

Display machine information:
```{r}
sessionInfo()
```
Display my machine memory.
```{r}
memuse::Sys.meminfo()
```

Load database libraries and the tidyverse frontend:
```{r}
library(bigrquery)
library(dbplyr)
library(DBI)
library(gt)
library(gtsummary)
library(tidyverse)
```

## Q1. Compile the ICU cohort in HW3 from the Google BigQuery database 

Below is an outline of steps. In this homework, we exclusively work with the BigQuery database and should not use any MIMIC data files stored on our local computer. Transform data as much as possible in BigQuery database and `collect()` the tibble **only at the end of Q1.7**.

### Q1.1 Connect to BigQuery

Authenticate with BigQuery using the service account token. Please place the service account token (shared via BruinLearn) in the working directory (same folder as your qmd file). Do **not** ever add this token to your Git repository. If you do so, you will lose 50 points.
```{r}
# path to the service account token 
satoken <- "biostat-203b-2025-winter-4e58ec6e5579.json"
# BigQuery authentication using service account
bq_auth(path = satoken)
```
Connect to BigQuery database `mimiciv_3_1` in GCP (Google Cloud Platform), using the project billing account `biostat-203b-2025-winter`.
```{r}
# connect to the BigQuery database `biostat-203b-2025-mimiciv_3_1`
con_bq <- dbConnect(
    bigrquery::bigquery(),
    project = "biostat-203b-2025-winter",
    dataset = "mimiciv_3_1",
    billing = "biostat-203b-2025-winter"
)
con_bq
```
List all tables in the `mimiciv_3_1` database.
```{r}
dbListTables(con_bq)
```

### Q1.2 `icustays` data

Connect to the `icustays` table.
```{r}
# full ICU stays table
icustays_tble <- tbl(con_bq, "icustays") |>
  arrange(subject_id, hadm_id, stay_id) |>
   # show_query() |>
  print(width = Inf)
```

### Q1.3 `admissions` data

Connect to the `admissions` table.
```{r}
# create admissions table from con_bq
 admissions_tble <- tbl(con_bq, "admissions") |>
  arrange(subject_id, hadm_id) |>
  print(width = Inf)

```

### Q1.4 `patients` data

Connect to the `patients` table.
```{r}
# creating patients table from con_bq
 patients_tble <- tbl(con_bq, "patients") |>
  arrange(subject_id) |>
  print(width = Inf) # unlimited columns printed

```

### Q1.5 `labevents` data

Connect to the `labevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the lab items listed in HW3. Only keep the last lab measurements (by `storetime`) before the ICU stay and pivot lab items to become variables/columns. Write all steps in _one_ chain of pipes.
```{r}
#filter store lab measurments
lab_item_ids <- c(50912, 50971, 50983, 50902, 
                  50882, 51221, 51301, 50931)

# create labevents table
 labevents_tble <- tbl(con_bq, "labevents") |>
   # we only want subject_id, item_id, storetime, valuenum
   select(subject_id, itemid, storetime, valuenum) |>
    # Join with icustays_tble on subject_id
  inner_join(icustays_tble, by = "subject_id") |>
   
     # Filter lab events that are before ICU stay
  filter(storetime < intime) |>
  
  select(subject_id, itemid, storetime, valuenum) |>
  
   group_by(subject_id, itemid) |>
   
   filter(storetime == max(storetime)) |>
   
  filter(itemid %in% lab_item_ids) |>
   
     # Pivot wider to make lab items into columns
  pivot_wider(names_from = itemid, values_from = valuenum) |>
   
  arrange(subject_id, storetime, itemid) |>
   
  print(width = Inf) # unlimited columns printed

```

### Q1.6 `chartevents` data

Connect to `chartevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the chart events listed in HW3. Only keep the first chart events (by `storetime`) during ICU stay and pivot chart events to become variables/columns. Write all steps in _one_ chain of pipes. Similary to HW3, if a vital has multiple measurements at the first `storetime`, average them.
```{r}

# Define chart item IDs from HW3
chart_item_ids <- c(220045, 220179, 220180, 223761, 220210)

# # Create chart events table
chartevents_tble <- tbl(con_bq, "chartevents") |>
  # Join with icustays_tble on subject_id
  inner_join(icustays_tble, by = "subject_id") |>
  
  # Filter chart events that occur during ICU stay (between intime and outtime)
  filter(storetime >= intime & storetime <= outtime) |>
  
  # Select relevant columns:
  select(subject_id, itemid, storetime, value) |>
  
  # Convert 'value' to numeric (handling any possible conversion errors with NA)
  mutate(value = as.numeric(value)) |>
  
  # Filter for chart events that are in the HW3 list
  filter(itemid %in% chart_item_ids) |>
  
  # Group by subject_id and itemid to handle cases 
  # where there are multiple measurements at the same time
  group_by(subject_id, itemid) |>
  
  # Keep only the first measurement for each subject 
  # and chart event (earliest storetime)
  filter(storetime == min(storetime)) |>
  
  # If multiple measurements exist at the first 
  # storetime, take the average of `value`
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop") |>
  
  # Pivot wider to turn itemid into columns and
  # each column will contain the corresponding value
  pivot_wider(names_from = itemid, values_from = value) |>
  
  # Arrange by subject_id
  arrange(subject_id) |>
  
  print(width = Inf)  # Unlimited columns printed
```

### Q1.7 Put things together

This step is similar to Q7 of HW3. Using _one_ chain of pipes `|>` to perform following data wrangling steps: (i) start with the `icustays_tble`, (ii) merge in admissions and patients tables, (iii) keep adults only (age at ICU intime >= 18), (iv) merge in the labevents and chartevents tables, (v) `collect` the tibble, (vi) sort `subject_id`, `hadm_id`, `stay_id` and `print(width = Inf)`.

```{r}
# aggregate labevents at hadm_id level before joining
labevents_tble <- tbl(con_bq, "labevents") |>  
  group_by(subject_id, hadm_id) |>  
  summarise(valuenum_avg = mean(valuenum, na.rm = TRUE), .groups = "drop")

# Aggregate chartevents at stay_id level before joining
chartevents_tble <- tbl(con_bq, "chartevents") |>  
  group_by(subject_id, stay_id) |>  
  summarise(valuenum_avg = mean(valuenum, na.rm = TRUE), .groups = "drop")

# create mimic cohort 
mimic_icu_cohort <- icustays_tble |>  
  # Merge with admissions table
  inner_join(tbl(con_bq, "admissions"), by = c("subject_id", "hadm_id")) |>  
  # Merge with patients table
  inner_join(tbl(con_bq, "patients"), by = "subject_id") |>  
  # Correct way to calculate age in MIMIC-IV
  mutate(age = as.integer(anchor_age + (year(intime) - anchor_year))) |>  
  filter(age >= 18) |>  # Keep only adults
  # Merge with summarized labevents (Avoids explosion of rows)
  inner_join(labevents_tble, by = c("subject_id", "hadm_id")) |>  
  # Merge with summarized chartevents (Avoids explosion of rows)
  inner_join(chartevents_tble, by = c("subject_id", "stay_id")) |>  
  # Collect at the very end (Prevents unnecessary memory usage)
  collect() |>  
  # Remove duplicates (Keeps only one row per subject_id + hadm_id + stay_id)
  distinct(subject_id, hadm_id, stay_id, .keep_all = TRUE) |>  
  # Sort the final dataset
  arrange(subject_id, hadm_id, stay_id) |>  
  print(width = Inf)

```

### Q1.8 Preprocessing

Perform the following preprocessing steps. (i) Lump infrequent levels into "Other" level for `first_careunit`, `last_careunit`, `admission_type`, `admission_location`, and `discharge_location`. (ii) Collapse the levels of `race` into `ASIAN`, `BLACK`, `HISPANIC`, `WHITE`, and `Other`. (iii) Create a new variable `los_long` that is `TRUE` when `los` is greater than or equal to 2 days. (iv) Summarize the data using `tbl_summary()`, stratified by `los_long`. Hint: `fct_lump_n` and `fct_collapse` from the `forcats` package are useful.

```{r}
# Load necessary packages
library(dplyr)
library(forcats)
library(gtsummary)


  # Step (i) Lump infrequent levels into "Other" level for specific columns
  # we take the top 5 most frequent values, rest are lumped into "other"
  mutate(mimic_icu_cohort,
         
         first_careunit = fct_lump_n(first_careunit, n = 5,
                                     other_level = "Other"),
         last_careunit = fct_lump_n(last_careunit, n = 5,
                                    other_level = "Other"),
         admission_type = fct_lump_n(admission_type, n = 5,
                                     other_level = "Other"),
         admission_location = fct_lump_n(admission_location, n = 5,
                                         other_level = "Other"),
         discharge_location = fct_lump_n(discharge_location, n = 5,
                                         other_level = "Other"),
  
    # Step (ii) Collapse the levels of race into specified categories
     race = fct_collapse(race,
                    ASIAN = c("ASIAN - SOUTH EAST ASIAN",
                              "ASIAN", "ASIAN - CHINESE", 
                              "ASIAN - KOREAN",
                              "ASIAN - ASIAN INDIAN"),
                    BLACK = c("BLACK/AFRICAN AMERICAN",
                              "BLACK/CAPE VERDEAN",
                              "BLACK/AFRICAN",
                              "BLACK/CARIBBEAN ISLAND"),
                    HISPANIC = c("HISPANIC/LATINO - SALVADORAN",
                                 "HISPANIC/LATINO - PUERTO RICAN", 
                                 "HISPANIC/LATINO - GUATEMALAN",
                                 "HISPANIC/LATINO - CUBAN", 
                                 "HISPANIC/LATINO - DOMINICAN",
                                 "HISPANIC/LATINO - CENTRAL AMERICAN", 
                                 "HISPANIC/LATINO - HONDURAN",
                                 "HISPANIC/LATINO - COLUMBIAN", 
                                 "HISPANIC/LATINO - MEXICAN",
                                 "HISPANIC OR LATINO"),
                    WHITE = c("WHITE", "WHITE - RUSSIAN",
                              "WHITE - OTHER EUROPEAN",
                              "WHITE - BRAZILIAN", 
                              "WHITE - EASTERN EUROPEAN"),
                    Other = c("OTHER", "UNKNOWN", "UNABLE TO OBTAIN",
                              "PATIENT DECLINED TO ANSWER", 
                              "SOUTH AMERICAN",
                              "MULTIPLE RACE/ETHNICITY", 
                              "AMERICAN INDIAN/ALASKA NATIVE",
                              "NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER")),
    
    # Step (iii) Create a new variable `los_long` based on `los`
    los_long = los >= 2
) |>
  
  # Step (iv) Summarize the data using tbl_summary(), stratified by los_long
  tbl_summary(by = c(los_long)) |>
  
  # Print the summary
  print()
```


### Q1.9 Save the final tibble

Save the final tibble to an R data file `mimic_icu_cohort.rds` in the `mimiciv_shiny` folder.
```{r}
# make a directory mimiciv_shiny
if (!dir.exists("mimiciv_shiny")) {
  dir.create("mimiciv_shiny")
}
# save the final tibble
mimic_icu_cohort |>
  write_rds("mimiciv_shiny/mimic_icu_cohort.rds", compress = "gz")
```
Close database connection and clear workspace.
```{r}
if (exists("con_bq")) {
  dbDisconnect(con_bq)
}
rm(list = ls())
```
Although it is not a good practice to add big data files to Git, for grading purpose, please add `mimic_icu_cohort.rds` to your Git repository.

## Q2. Shiny app

Develop a Shiny app for exploring the ICU cohort data created in Q1. The app should reside in the `mimiciv_shiny` folder. The app should contain at least two tabs. One tab provides easy access to the graphical and numerical summaries of variables (demographics, lab measurements, vitals) in the ICU cohort, using the `mimic_icu_cohort.rds` you curated in Q1. The other tab allows user to choose a specific patient in the cohort and display the patient's ADT and ICU stay information as we did in Q1 of HW3, by dynamically retrieving the patient's ADT and ICU stay information from BigQuery database. Again, do **not** ever add the BigQuery token to your Git repository. If you do so, you will lose 50 points.


```{r}
#installs

#install.packages("shinydashboard")
#install.packages('DT')
#install.packages("gt")
# Load necessary libraries
library(shiny)
library(shinydashboard)
library(dplyr)
library(ggplot2)
library(DT)
library(gt)
library(gtsummary)
library(bigrquery)

# Load the mimic_icu_cohort data (assuming it's saved as .rds file)
mimic_icu_cohort <- readRDS("mimiciv_shiny/mimic_icu_cohort.rds")

# UI of the app
ui <- dashboardPage(
  dashboardHeader(title = "ICU Cohort Data Exploration"),
  
  dashboardSidebar(
    sidebarMenu(
      menuItem("Summary", tabName = "summary", icon = icon("dashboard")),
      menuItem("Patient Info", tabName = "patient_info", icon = icon("user"))
    )
  ),
  
  dashboardBody(
    tabItems(
      
      # Tab 1: Summary Statistics
      tabItem(tabName = "summary",
              fluidRow(
                box(title = "Select Variable for Bar Plot",
                    width = 4, solidHeader = TRUE, status = "info",
                    selectInput("barplot_var", "Choose Variable for Bar Plot",
                                choices = c("Insurance" = "insurance",
                                            "Marital Status" = "marital_status",
                                            "Race" = "race",
                                            "Gender" = "gender"
                                            )
                                )
                  ),
                box(title = "Bar Plot Summary",
                    width = 8, solidHeader = TRUE, status = "primary", 
                    plotOutput("barplot_summary")
                    )
                      )
              ),
      
      # Tab 2: Patient Information
      tabItem(tabName = "patient_info",
        fluidRow(
          box(title = "Enter Patient ID",
              width = 4, solidHeader = TRUE, status = "primary",
              selectInput("patient_id",
                          "Choose Patient ID",
                          choices = unique(mimic_icu_cohort$subject_id))
            ),
          box(title = "Patient ADT",
              width = 8, solidHeader = TRUE, status = "primary",
              plotOutput("patient_plot"))  # Render the plot here
                )
              )
          )
            )
)


# Server logic
server <- function(input, output, session) {
  
  # Tab 1: Numerical Summaries using gtsummary
  output$num_summary <- render_gt({
    # Summarize the data stratified by the `los_long` variable
    mimic_icu_cohort %>%
      tbl_summary(by = los_long) %>%
      as_gt() %>% 
      datatable()
  })
  
  # Tab 1: Bar Plot Summary based on selected variable
  output$barplot_summary <- renderPlot({
    req(input$barplot_var)  # Ensure that a variable is selected
    
    # Select the appropriate variable based on input
    var_name <- input$barplot_var
    plot_data <- mimic_icu_cohort %>%
      count(!!sym(var_name)) %>%
      rename(Variable = !!sym(var_name), Count = n)
    
    # Generate the bar plot
    ggplot(plot_data, aes(x = Variable, y = Count, fill = Variable)) +
      geom_bar(stat = "identity", color = "black") + # Unique fill color
      # print count above bar plot
      geom_text(aes(label = Count), vjust = -0.5, size = 4) +  
      theme_minimal() +
      labs(title = paste(var_name, "Summary"), x = var_name, y = "Count") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  })
  
  # Filter the data for the selected patient ID
  patient_data_filtered <- reactive({
    req(input$patient_id)  # Ensure that a patient_id is selected
    mimic_icu_cohort %>% filter(subject_id == input$patient_id)
  })
  
   # Get patient demographic info (for title)
  patient_info <- reactive({
    req(input$patient_id)
    patient_data <- patient_data_filtered()
    # Extract patient demographics
    demographic_info <- patient_data %>%
      summarise(
        gender = first(gender),
        age = first(age),
        race = first(race),
        insurance = first(insurance)
      )
    return(demographic_info)
  })
  
  
  # Tab 2: Render the ADT or ICU plot based on user selection
  output$patient_plot <- renderPlot({
    
    # Filter the data for the selected patient
    patient_data <- patient_data_filtered()
    
    # Generate the ADT plot
    
    plot_title <- paste("Patient", input$patient_id, ",",
                        patient_info()$gender, ",",
                        patient_info()$age, "years old,",
                        patient_info()$race)
    
      ggplot(patient_data, aes(x = admittime, y = admission_type)) +
      geom_segment(aes(xend = dischtime, yend = admission_type,
                       color = admission_location, size = 4)) +
      geom_point(aes(x = admittime, y = admission_type, shape = "ADT"),
                 size = 4) +
      geom_point(aes(x = dischtime, y = admission_type, shape = "ADT"),
                 size = 4) +
      scale_color_manual(values = c("ICU" = "red", "CCU" = "blue",
                                    "Other" = "gray"),
                         name = "Care Unit") +
      scale_size_continuous(range = c(1, 3)) +
        # Circle shape for ADT events
      scale_shape_manual(values = c("ADT" = 16)) +  
      labs(title = plot_title,
           x = "Calendar Time", y = "Event Type",
           color = "Care Unit", size = "ICU/CCU Size") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    }
)}

# Run the app
shinyApp(ui, server)

```
```

