---
title: "Biostat 203B Homework 5"
subtitle: Due Mar 20 @ 11:59PM
author: "Nathaniel Boyle and 006525153"
format:
  html:
    theme: cosmogit
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

## Predicting ICU duration

Using the ICU cohort `mimiciv_icu_cohort.rds` you built in Homework 4, develop
at least three machine learning approaches (logistic regression with enet
regularization, random forest, boosting, SVM, MLP, etc) plus a model 
stacking approach for predicting whether a patient's ICU stay will be 
longer than 2 days. You should use the `los_long` variable as the outcome. 
You algorithms can use patient demographic information 
(gender, age at ICU `intime`, marital status, race),
ICU admission information (first care unit),
the last lab measurements before the ICU stay, and first vital measurements
during ICU stay as features. 
You are welcome to use any feature engineering techniques you think are
appropriate; but make sure to not use features that are not available at an ICU
stay's `intime`. For instance, `last_careunit` cannot be used in your
algorithms. 

1. Data preprocessing and feature engineering.

Following slides on logistic regression

```{r}
#Display system infomration for reproducibility
sessionInfo()
```
```{r setup, include=FALSE}
#ignore warnings in html output
knitr::opts_chunk$set(warning = FALSE)
```

```{r}
# Installs
# Check if tidymodels is installed, and install if not
# Check if tidymodels, rlang, or cli are installed, and install them if not

if (!requireNamespace("tidymodels", quietly = TRUE)) {
  install.packages("tidymodels")
}

# Check if rlang is installed, and install it if not
if (!requireNamespace("rlang", quietly = TRUE)) {
  install.packages("rlang")
}

# Check if cli is installed, and install it if not
if (!requireNamespace("cli", quietly = TRUE)) {
  install.packages("cli")
}
# Check if cli is installed, and install it if not
if (!requireNamespace("ranger", quietly = TRUE)) {
  install.packages("ranger")
}
# Check if cli is installed, and install it if not
if (!requireNamespace("keras", quietly = TRUE)) {
  install.packages("keras")
}
# Check if cli is installed, and install it if not
if (!requireNamespace("stacks", quietly = TRUE)) {
  install.packages("stacks")
}

# Load libraries
library(GGally)
library(gtsummary)
library(tidyverse)
library(tidymodels)
library(dplyr)
library(forcats)
library(gtsummary)
library(ranger)
library(keras)
library(stacks)
```

```{r}
version
```

```{r}
# Load the mimic_icu_cohort data 

mimiciv_icu_cohort <- readRDS("mimic_icu_cohort.rds") |>
  
 # Remove columns not needed for analysis
   select(-admit_provider_id, admittime, dischtime , deathtime, edregtime, 
          edouttime) |>
  
  # Lump infrequent levels into "Other" level for specific columns
  # we take the top 5 most frequent values, rest are lumped into "other"
  mutate(first_careunit = fct_lump_n(first_careunit, n = 5,
                                     other_level = "Other"),
         last_careunit = fct_lump_n(last_careunit, n = 5,
                                    other_level = "Other"),
         admission_type = fct_lump_n(admission_type, n = 5,
                                     other_level = "Other"),
         admission_location = fct_lump_n(admission_location, n = 5,
                                         other_level = "Other"),
         discharge_location = fct_lump_n(discharge_location, n = 5,
                                         other_level = "Other"),
  
    #  Collapse the levels of race into specified categories
     race = fct_collapse(race,
                    ASIAN = c("ASIAN - SOUTH EAST ASIAN",
                              "ASIAN", "ASIAN - CHINESE", 
                              "ASIAN - KOREAN",
                              "ASIAN - ASIAN INDIAN"),
                    BLACK = c("BLACK/AFRICAN AMERICAN",
                              "BLACK/CAPE VERDEAN",
                              "BLACK/AFRICAN",
                              "BLACK/CARIBBEAN ISLAND"),
                    HISPANIC = c("HISPANIC/LATINO - SALVADORAN",
                                 "HISPANIC/LATINO - PUERTO RICAN", 
                                 "HISPANIC/LATINO - GUATEMALAN",
                                 "HISPANIC/LATINO - CUBAN", 
                                 "HISPANIC/LATINO - DOMINICAN",
                                 "HISPANIC/LATINO - CENTRAL AMERICAN", 
                                 "HISPANIC/LATINO - HONDURAN",
                                 "HISPANIC/LATINO - COLUMBIAN", 
                                 "HISPANIC/LATINO - MEXICAN",
                                 "HISPANIC OR LATINO"),
                    WHITE = c("WHITE", "WHITE - RUSSIAN",
                              "WHITE - OTHER EUROPEAN",
                              "WHITE - BRAZILIAN", 
                              "WHITE - EASTERN EUROPEAN"),
                    Other = c("OTHER", "UNKNOWN", "UNABLE TO OBTAIN",
                              "PATIENT DECLINED TO ANSWER", 
                              "SOUTH AMERICAN",
                              "MULTIPLE RACE/ETHNICITY", 
                              "AMERICAN INDIAN/ALASKA NATIVE",
                              "NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER")),
    
    # Create a new variable `los_long` based on `los`
    los_long = los >= 2
)

# create summary (stratified by los_long)
summary_table <- mimiciv_icu_cohort |>
  tbl_summary(by = c("los_long"))

# Print the summary table
print(summary_table)

```

2. Partition data into 50% training set and 50% test set. Stratify partitioning according to `los_long`. For grading purpose, sort the data by `subject_id`, `hadm_id`, and `stay_id` and use the seed `203` for the initial data split. Below is the sample code.
```{r}
#| eval: false
set.seed(203)

#Sort the data by subject_id, hadm_id, and stay_id
mimiciv_icu_cohort <- mimiciv_icu_cohort |>
  arrange(subject_id, hadm_id, stay_id) |>
  select(-c(intime, outtime, admittime, dischtime, deathtime, edregtime,
            edouttime, dod, subject_id, hadm_id, stay_id, anchor_age))

# Perform the initial split (stratifying by los_long)
data_split <- initial_split(
  mimiciv_icu_cohort,
  strata = "los_long",
  prop = 0.5
)
```

```{r}
#check training output
mimiciv_other <- training(data_split)
dim(mimiciv_other)
```

```{r}
#check test output
mimiciv_test <- training(data_split)
dim(mimiciv_test)
```


3. Train and tune the models using the training set.

```{r}
# Logistic Regression recipe

# 

logit_recipe <- recipe(
    los_long ~ .,  # Stratify on los_long
    data = mimiciv_other
  ) |>
  # Convert 'los_long' to factor (if it isn't already a factor)
  step_mutate(los_long = as.factor(los_long)) |>
  # Impute missing values for numeric columns (mean imputation)
  step_impute_mean(all_numeric_predictors()) |>
  # Impute missing values for categorical/logical columns (mode imputation)
  step_impute_mode(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  # Create dummy variables for nominal predictors
  step_dummy(all_nominal_predictors()) |>
  # Zero-variance filter for numeric predictors
  step_zv(all_numeric_predictors()) |>
  # Normalize numeric predictors
  step_normalize(all_numeric_predictors())

# Optionally, you can inspect the recipe
logit_recipe |> print()


# # Check if there are any remaining missing values in the dataset
 anyNA(mimiciv_other)
 
 table(mimiciv_other$los_long)

```

```{r}
# Logistic Regression Model

logit_mod <- 
  logistic_reg(
    penalty = tune(), 
    mixture = tune()
  ) |> 
  set_engine("glmnet", standardize = FALSE) |>
  print()

```
```{r}
#Set Logistic Workflow
logit_wf <- workflow() |>
  add_recipe(logit_recipe) |>
  add_model(logit_mod) |>
  print()
```

```{r}
# Create grid, tune penatly and mixure hyperparameters
param_grid <- grid_regular(
  penalty(range = c(-6, 3)), 
  mixture(),
  levels = c(100, 5)
  ) |>
  print()
```

```{r}
# Cross Validation Set up
folds <- vfold_cv(mimiciv_other, v = 5)
folds

```
```{r}
#fit cross-validation

(logit_fit <- logit_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )) |>
  system.time()

```


```{r}
logit_fit
```


```{r}
#vizualize Cross Validation results

logit_fit |>
  # aggregate metrics from K folds
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = penalty, y = mean, color = factor(mixture))) +
  geom_point() +
  labs(x = "Penalty", y = "CV AUC") +
  scale_x_log10()

```


Training SVM Model

```{r}

svm_recipe <- 
  recipe(
    AHD ~ ., 
    data = Heart_other
  ) |>
  # mean imputation for Ca
  step_impute_mean(Ca) |>
  # mode imputation for Thal
  step_impute_mode(Thal) |>
  # create traditional dummy variables (necessary for svm)
  step_dummy(all_nominal_predictors()) |>
  # zero-variance filter
  step_zv(all_numeric_predictors()) |> 
  # center and scale numeric data
  step_normalize(all_numeric_predictors()) |>
  # estimate the means and standard deviations
  # prep(training = Heart_other, retain = TRUE) |>
  print()


svm_mod <- 
  svm_rbf(
    mode = "classification",
    cost = tune(),
    rbf_sigma = tune()
  ) |> 
  set_engine("kernlab")
svm_mod


svm_wf <- workflow() |>
  add_recipe(svm_recipe) |>
  add_model(svm_mod)
svm_wf


param_grid <- grid_regular(
  cost(range = c(-8, 5)),
  rbf_sigma(range = c(-5, -3)),
  levels = c(14, 5)
  )
param_grid


set.seed(203)

folds <- vfold_cv(Heart_other, v = 5)
folds


svm_fit <- svm_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
svm_fit


svm_fit <- svm_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
svm_fit


svm_fit |>
  show_best(metric = "roc_auc")


final_wf <- svm_wf |>
  finalize_workflow(best_svm)
final_wf


# Fit the whole training set, then predict the test cases
final_fit <- 
  final_wf |>
  last_fit(data_split)
final_fit


```

Train Random Forest Model

```{r}

rf_recipe <- 
  recipe(
    AHD ~ ., 
    data = Heart_other
  ) |>
  # # create traditional dummy variables (not necessary for random forest in R)
  # step_dummy(all_nominal()) |>
  # mean imputation for Ca
  step_impute_mean(Ca) |>
  # mode imputation for Thal
  step_impute_mode(Thal) |>
  # zero-variance filter
  step_zv(all_numeric_predictors()) |>
  # # center and scale numeric data (not necessary for random forest)
  # step_normalize(all_numeric_predictors()) |>
  # estimate the means and standard deviations
  # prep(training = Heart_other, retain = TRUE) |>
  print()


rf_mod <- 
  rand_forest(
    mode = "classification",
    # Number of predictors randomly sampled in each split
    mtry = tune(),
    # Number of trees in ensemble
    trees = tune()
  ) |> 
  set_engine("ranger")
rf_mod

rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_mod)
rf_wf


param_grid <- grid_regular(
  trees(range = c(100L, 500L)), 
  mtry(range = c(1L, 5L)),
  levels = c(5, 5)
  )
param_grid

set.seed(203)

folds <- vfold_cv(Heart_other, v = 5)
folds

rf_fit <- rf_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
rf_fit

```
Stacking the 3 models


```{r}
heart_model_st <- 
  # initialize the stack
  stacks() |>
  # add candidate members
  add_candidates(logit_res) |>
  add_candidates(rf_res) |>
  add_candidates(mlp_res) |>
  # determine how to combine their predictions
  blend_predictions(
    penalty = 10^(-6:2),
    metrics = c("roc_auc")
    ) |>
  # fit the candidates with nonzero stacking coefficients
  fit_members()
heart_model_st

autoplot(heart_model_st)

autoplot(heart_model_st, type = "members")

autoplot(heart_model_st, type = "weights")


collect_parameters(heart_model_st, "rf_res")


heart_pred <- Heart_test %>%
  bind_cols(predict(heart_model_st, ., type = "prob")) %>%
  print(width = Inf)


yardstick::roc_auc(
  heart_pred,
  truth = AHD,
  contains(".pred_No")
  )


heart_pred <-
  Heart_test |>
  select(AHD) |>
  bind_cols(
    predict(
      heart_model_st,
      Heart_test,
      type = "class",
      members = TRUE
      )
    ) |>
  print(width = Inf)


map(
  colnames(heart_pred),
  ~mean(heart_pred$AHD == pull(heart_pred, .x))
  ) |>
  set_names(colnames(heart_pred)) |>
  as_tibble() |>
  pivot_longer(c(everything(), -AHD))
```


4. Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each machine learning algorithm and the model stacking. Interpret the results. What are the most important features in predicting long ICU stays? How do the models compare in terms of performance and interpretability?
